Effects of pulsed hollow electron lens operation on the beam core in HL-LHC: \\First experimental studies and simulations.

### 2018-02-06 resex.tex
Changes Stefano, 2018-02-06

\begin{abstract}
In case of the High Luminosity Large Hadron Collider (HL-LHC), a considerable amount of energy is stored in the beam tails due to the high beam intensity and, in addition, an overpopulation of the tails compared to a Gaussian distribution. To control and clean the tail population, the installation of two hollow electron lenses, one per beam, is foreseen. Beside the DC operation of the lenses, also pulsed operation is considered having the advantage to considerably increase the diffusion speed by exerting noise on the halo particles. In the ideal case, that is in case of no field at the proton beam core, only the halo particles are excited while leaving the core unperturbed. The picture though changes, if a residual field is present also at the location of the beam core exposing also the the particles in the center to noise. In this paper, we present first estimates of the residual field at the proton beam core expected from the HL-LHC hollow electron lens and summarize the results of the LHC experiments conducted in 2016 and 2017 to study the effects on the beam core in case of pulsed operation.
\end{abstract}

@@

\section{Introduction\label{sec:intro}}%force line break with \protect\\
Considering past, current and future high energy and intensity colliders, each new machine has represented a considerable leap in stored beam energy with rising values for future accelerators and colliders (see Table~\ref{tab:stored_energy}). 
\begin{table*}
	\caption{\label{tab:stored_energy}%
		Stored beam energy for different past, present and future colliders. Each new machine represents a leap in stored beam energy.
	}
	\begin{ruledtabular}
		\begin{tabular}{lccccc}
			Collider& Tevatron (protons) \cite{tevatron} & LHC 2016 \cite{chamonix2017param}
			& LHC nominal \cite{lhc_design} & HL-LHC \cite{hlcdr} & FCC \cite{fcc_param_2017} \\
			\colrule
			Beam energy [TeV] & 0.98 & 6.5 & 7.0 & 7.0 & 50.0\\
			Number of bunches & 36 & 2220 & 2808 & 2748 & ? \\
			Number of particles per bunch & $2.90\times 10^{11}$ & $1.15\times 10^{11}$ & $1.15\times 10^{11}$ & $2.2\times 10^{11}$ & $1.0\times 10^{11}$\\
			Stored beam energy [MJ] & 1.6 & 265.9 & 362.2 & 678.0 & 8400 \\
		\end{tabular}
	\end{ruledtabular}
\end{table*}
Recent measurements at the LHC furthermore show that the tails of the transverse beam distribution are overpopulated compared to a Gaussian distribution resulting in a considerable amount of energy being stored in the beam tails alone. In case of the LHC explicitly around 5\% of the beam population is stored in the tails above 3.5~$\sigma$ compared to 0.22\% in case of a Gaussian distribution leading to 19~MJ of stored energy in the tails in case of nominal LHC parameters and 34~MJ in case of HL-LHC~\cite{helreview_valentino}.

All of the above reasons lead to the conclusion that a mechanism is needed to deplete the beam tails in a controlled manner \cite{helreview}. The most obvious idea is to decrease the collimator gaps or scrape the tails with a collimator type device. The minimum gap size of the collimators is however limited by the impedance of the machine. Small enough gap sizes can therefore not be reached without risking severe beam instabilities. Other mechanisms must therefore be found to actively deplete the tails. Most promising are methods, which increase the diffusion speed in the region of the halo particles while leaving the beam core unperturbed. The diffusing halo particles are then intercepted by the collimation system and removed. This concept is also referred to as active halo control compared to a passive system, which only intercepts the halo particles without actively controlling the diffusion speed (see Fig.~\ref{fig:active_halo_control} for illustration).
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.7\linewidth]{passive_halo_control.png}
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.7\linewidth]{active_halo_control.png}
	\end{minipage}	
	\caption{\label{fig:active_halo_control} Sketch of passive halo control as with the collimation system (top) and active halo control using in addition for example a hollow electron lens (HEL) to control the diffusion speed in the region tail region without affected the beam core (bottom).}
\end{figure}

In a recent review, the need for such an active halo control system for HL-LHC has been assessed with the conclusion that it would considerably increase the margin and reduce the risk for machine protection compared to the current LHC collimation system being a passive halo control system ~\cite{helreview}. In view of the need of active halo control for HL-LHC and for future high power accelerators like HE-LHC and FCC-hh~\cite{helhcparam2011,fcc_coll_ipac2017}, different active halo control methods have been studied in recent years~\cite{helreview_bruce}, of which the hollow electron lens (HEL) is considered the superior device for the HL-LHC~\cite{helreview}. The beneficial effect of a future HEL for HL-LHC in terms of machine protection would be however abrogated if the beam core was negatively affected leading ultimately to a degradation of the performance due to increased particle losses and emittance growth. In this paper, we will concentrate on this aspect. We will summarize possible sources of perturbations of the proton beam core focusing in particular on the case of pulsed operation and the beam experiments at the LHC conducted in this context.

@@

%This includes the detailed analysis of the observed losses, emittance and transverse beam distribution changes. To the knowledge of the authors, the observed distribution changes presented in this paper have never been measured before in the context of a resonant excitation. A resonant excitation has been previously studied at the Tevatron in the context of the HEL experiments \cite{hel_tevatron_stancari} and the abort gap cleaning used in operation \cite{hel_tevatron_abortgap_zhang}. Both studies however only concentrated on the losses and emittance changes without measuring the detailed changes of the distribution itself. In addition, the presented experiments also provide scalings of the losses and emittance growth with the excitation amplitude. This allows to provide first tolerances for the residual field of the HL-LHC HEL at the beam core and also to compare the obtained results with simulations and thus benchmark the same.

\section{Hollow electron lens for HL-LHC\label{sec:hel}}
\subsection{General Overview\label{sec:hel:intro}}
Electron lenses are in general DC or pulsed low energy electron beams, where the electron beam is generated in an electron gun, then guided and confined by strong solenoids and finally dumped onto a collector. As an example the conceptual design of the HL-LHC HEL is shown in Fig.~\ref{fig:hel_layout}.
\begin{figure}[h]
	\includegraphics[width=1.0\linewidth]{hel_layout_epbeam}% Here is how to import EPS art
	\caption{\label{fig:hel_layout} Layout of HL-LHC HEL.}
\end{figure}

@@

			Desired range of scraping positions\footnote{$\sigma$ denotes here the Beam sigma and not the collimation $\sigma$ which assumes a normalized emittance of 3.5~$\mu$m.} & 4-8 &$\sigma_p$\\

@@

\subsection{Operation modes and effects on the beam core\label{sec:hel:core}}
As depicted in Fig.~\ref{fig:hel_field} and in evidence from Eqs.~\ref{eq:field_1}--\ref{eq:field_2}, the field at the beam core vanishes in case of an ideal HEL. Effects on the beam core thus only arise from imperfections, where possible sources are the bends of the HEL (see Section~\ref{core:sec:1}), as here the electron beam crosses directly the proton beam, and distortions in the electron beam profile (see Section~\ref{core:sec:2}). Both sources result in non-linear kicks~\cite{hel_bends_stancari,hel_model_polynomial_morozov}. In DC operation, these non-linear kicks stay in the shadow of the non-linearities otherwise present in the machine. Tolerances on imperfections are therefore not particularly stringent and of no concern. The picture changes significantly in case of pulsed operation, in which case the electron gun voltage is modulated using a white or colored noise spectrum. If the electromagnetic field of the HEL does not vanish at the proton beam core, noise is transferred not only to the halo particles, as intended, but also to the beam core. Tolerances on the residual field in this case rapidly become much more stringent than in the case of DC operation. Studies of the effect of the HEL on the beam core are therefore focusing on this mode of operation which will also be the main subject of this paper.

The main benefit of pulsed HEL operation is the increase in halo removal rate particularly relevant if a fast depletion of the halo is desired. The halo diffusion rate increases in this case as in addition to the non-linear kick, noise is exerted on the halo particles. For the HL-LHC two different pulsing patterns are considered for this purpose:
\begin{itemize}
	\item \textbf{random excitation:} The electron gun voltage is modulated between:
	\begin{eqnarray}
		U_{\mathrm{e-gun}}&=&a\cdot U_{\mathrm{max}}\\
		& &+(1-a)\cdot \mathrm{ran}(0,1)\cdot U_{\mathrm{max}}
	\end{eqnarray}
	with $U_{\mathrm{max}}$ the equivalent voltage in DC operation, $a$ the modulation strength with $a\in[0,1]$, and $\mathrm{ran}(0,1)$ a uniformly distributed random number between~[0,1].
	\item \textbf{resonant excitation:} The HEL is switched on only every $k^{\mathrm{th}}$ turn. The excitation can then be represented by:
	\begin{eqnarray}\label{intro:eqn:1}
	f(t)&=&\sum_{p=-\infty}^{+\infty}\delta(t-n\cdot(kT)),
	\end{eqnarray}
	where $p$ is the turn number and $T$ the revolution time, and its Fourier series by:
	\begin{eqnarray}\label{intro:eqn:2}
	f(t)=\Sh_{kT}(t)
	&=&\frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt} \\
	& & \quad \text{with} \ f_n=\frac{n}{k}f_{\rm rev},
	\end{eqnarray}
	where $\Sh_{kT}$ is the Dirac comb. As can be seen from the Fourier series, $k^{\mathrm{th}}$~turn pulsing in general drives $k^{\mathrm{th}}$ order resonances \cite{md_sim_hel_res_ex_fitterer}. Historically, the $k^{\mathrm{th}}$~turn pulsing was used in regular operation in the Tevatron for abort cleaning \cite{hel_tevatron_abortgap_zhang}.
\end{itemize}
In general, the random excitation mode is the most effective and operationally easiest pulsing mode, but also the most stringent in terms of tolerances on the residual field at the proton beam center as it induces white noise exciting all beam frequencies. As the resonant excitation acts on specific resonances, the pulsing pattern has to be chosen according to the machine and beam parameters. On the one hand, this makes the operation more difficult as one has to find the most efficient $k^{\mathrm{th}}$~turn pulsing for each operation mode, on the other hand, it also relaxes the tolerances on the residual field at the beam core as, depending on the machine and beam configuration, it might be possible to find a resonant excitation pattern, which mainly affects the halo and not the beam core.

@@

\subsection{Uncompensated kicks from HEL bends}
\label{core:sec:1}
As reference for the estimate of the dipole component originating from the HEL bends, the symplectic map derived in \cite{hel_bends_stancari} is used. In this case the HEL bends are modeled as a cylindrical pipe with a static charge distribution of 1~A and 5~keV electrons deflected in the horizontal plane. In case of an U-shaped HEL, the transverse dipole kicks at entrance and exit add up, while for a S-shape the kicks compensate each other. For this reason a S-shape has been chosen for the HL-LHC HEL (see Fig.~\ref{fig:hel_layout}). Assuming and S-shaped HEL, uncompensated kicks therefore only arise from imperfections, which can originate from profile imperfections and current fluctuations. As a first estimate, we assume in this paper 10\% fluctuations between the entrance and exit and, in addition, that the kicks from entrance and exit due to these imperfections add up. Using the electric field calculations in \cite{hel_bends_stancari}, the maximum values for the integrated electric field are then
\begin{equation}
\int_{z_1}^{z_2} E_{x,y} dz= 10 \ \mathrm{kV}
\end{equation}
given an \SI{1}{A}, \SI{5}{keV} electron beam.
%For \SI{7}{TeV} protons and neglecting magnetic effects, this yields a kick of:
%\begin{equation}
%\Delta x'=\Delta y'= \SI{1.4}{nrad},
%\end{equation}
%Scaling to the HEL design parameters of a \SI{5}{A} and \SI{10}{keV} electron beam (Table~\ref{tab:hel_param}), one obtains:
%\begin{equation}
%\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
%\end{equation}
Neglecting magnetic effects and scaling to HL-LHC and HEL design parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}) yields a kick of:
\begin{equation}
\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
\end{equation}
Assuming now 10\% fluctuation between entrance and exit kick, the expected kick from the HEL bends is
\begin{equation}\label{eqn:kick_bends}
\Delta x', \Delta y'\approx \q{0.5}{nrad}.
\end{equation}

@@

\subsection{Kicks in the central region (main solenoid)}
\label{core:sec:2}
For a perfectly uniform, annular and radially symmetric profile, the field in the region of the proton beam core vanishes as can be seen from Eq.~\ref{eq:field_2} and is illustrated in Fig.~\ref{fig:hel_field}. Fields at the location of the proton beam core thus only arise in case of profile imperfections of the electron beam, in particular if the radial symmetry is broken. Recently, new measurements of the newest hollow electron gun CHG1b could be performed at the Fermilab electron lens test stand \cite{hel_test_stand_fnal}. A typical measurement of the electron beam current density obtained during these measurements is shown in Fig.~\ref{core:fig:0}.
\begin{figure}[t]
		\centering
		\includegraphics[width=0.8\linewidth]{e-field-measured.png}
	\caption{Current density from measurements of the CHG1b hollow electron gun at the Fermilab electron lens test stand in 2017 \cite{hel_res_field_stancari_2017}.}
	\label{core:fig:0}
\end{figure}
As at the test stand the solenoidal fields required for the HL-LHC electron lens can not be reached, an estimate of the electron beam profile for HL-LHC can only be obtained through simulations. For this purpose, the measured distribution is compressed to the inner electron beam radius of $0.97$~mm. Afterwards an input distribution with $40\ 000$ particles generated. The potential and fields can then be calculated with the particle in cell simulation code WARP~\cite{warp} yielding the relative electric field strengths shown in Fig.~\ref{core:fig:1}. The electric field can in general be scaled with the beam current and energy, but can not easily be scaled with the inner and outer radius of the electron beam distribution. Therefore it is essential to track a distribution with the correct dimensions. Further details on the measurements and WARP simulations can be found in~\cite{hel_res_field_stancari_2017}.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Emap.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Eslice.png}
	\end{minipage}
	\caption{Calculated relative electric field for the new CHG1b hollow electron gun in the x-y plane and as projection on the x- and y-axis. The field calculations are based on measurements at the Fermilab test stand combined with WARP simulations for the calculation of the electric potentials and fields.}
	\label{core:fig:1}
\end{figure}

@@

\subsection{Overview of LHC experiments\label{sec:exp_sum}}
The experiments at the LHC which are presented in this paper were aiming at deriving tolerances on the residual HEL field at the location of the beam core in case of pulsed operation. In total, two experiments were conducted, the first one in 2016 \cite{resexmd2016} and the second one in 2017 \cite{resexmd2017}. Beam and machine parameters for both experiments are summarized in Table~\ref{tab:md_param}. As there is currently no HEL installed in the LHC, the kick on the beam core was emulated by a dipole kick applied with the transverse damper of which more details will be given in Section~\ref{sec:adt}. Preceding the beam experiments, simulations were performed to identify the most effective $k^{\mathrm{th}}$~turn pulsing pattern and yield estimates for losses and emittance growth. The two $k^{\mathrm{th}}$~turn pulsing patterns with the largest effect, one $k^{\mathrm{th}}$~turn pulsing pattern with no effect and the random excitation could then be experimentally tested in the LHC. In order to test the reproducibility of the results at least on the example of one pulsing pattern, the $7^{\mathrm{th}}$~turn pulsing was first tried in 2016 and then repeated in 2017.

During the experiments the losses were measured with the fast beam current transformers (FBCTs) and the emittance and transverse beam profiles with the beam synchrotron radiation telescope (BSRT). The analysis of the BSRT profiles is quite involved and we will therefore focus in this paper only on the direct comparison of the profiles. A detailed description of the analysis with more details can be found in \cite{bsrtprofinj} and for the individual experiments in \cite{resexmd2016,resexmd2017}.

\subsection{Excitation with transverse damper and filling scheme\label{sec:adt}}
The primary function of the transverse feedback system in LHC, also referenced to as ADT, is to provide injection oscillation damping and actively damp the coupled bunch instabilities driven by the machine impedance. The main building blocks of the ADT system are stipline pickups at the position Q7 and Q9 at Point~4 of the LHC, connected to the beam position measurement modules at the surface, the digital signal processing modules (mDSPU) and a set of tetrode power amplifiers feeding electrostatic kickers in the RF zone of the LHC~\cite{adt_sum_2008,adt_sum_2011}. Though originally designed to damp the undesired transverse activities only, the damper is being routinely used for sophisticated beam excitations as well thanks to the state of the art digital signal processing hardware. These include among others for example abort and injection gap cleaning, single bunch excitation for tune and linear coupling measurements and special modes of operation for dedicated experiments at the LHC.

The transverse feedback is in general active during the whole LHC cycle, therefore it is of very high importance, that the noise introduced by the system does not contribute to any emittance growth. The typical machine cycle requires a short damping time of ~10-20 turns (high feedback gain) for injection oscillation damping, which is increased during the ramp and stable beams to ~50-100 turns (lower feedback gain). It has been demonstrated that noise introduced by the transverse feedback system itself is sufficiently low to introduce any measurable emittance growth \cite{adt_noise_emit_2017}. All excitation signals referenced to in this paper, are digitally synthesized in the damper's digital signal processing units and therefore can be considered as ``noise-free''. This point is important as then the effect on the beam can be attributed entirely to the applied excitation and any effect due to unwanted residual noise can be excluded.

>>>>

\begin{abstract}
At the High Luminosity Large Hadron Collider (HL-LHC), a considerable amount of energy is expected to be stored in the beam tails due to the high beam intensity and, in addition, an overpopulation of the tails compared to a Gaussian distribution. To control and clean the tail population, the installation of two hollow electron lenses, one per beam, is presently under consideration. Beside the DC operation of the lenses, also pulsed operation is considered having the advantage to considerably increase the diffusion speed by exerting white or colored noise on the halo particles. In an ideal electron beam with azimuthal symmetry and no perturbations at the incoming and outgoing electron beam locations, only the halo particles are excited while leaving the core unperturbed. The picture though changes, if a residual field is present also at the location of the beam core exposing also the the particles in the center to noise. In this paper, we present first estimates of the residual field at the proton beam core expected from the HL-LHC hollow electron lens and summarize the results of the LHC experiments conducted in 2016 and 2017 to study the effects on the beam core in case of pulsed operation.
\end{abstract}

@@

\section{Introduction\label{sec:intro}}%force line break with \protect\\
Considering past, current and future high energy and intensity colliders, each new machine has represented a considerable leap in stored beam energy with rising values for future accelerators and colliders (see Table~\ref{tab:stored_energy}). 
\begin{table*}
	\caption{\label{tab:stored_energy}%
		Stored beam energy for different past, present and future colliders. Each new machine represents a leap in stored beam energy.
	}
	\begin{ruledtabular}
		\begin{tabular}{lccccc}
			Collider& Tevatron (protons) \cite{tevatron} & LHC 2016 \cite{chamonix2017param}
			& LHC nominal \cite{lhc_design} & HL-LHC \cite{hlcdr} & FCC \cite{fcc_param_2017} \\
			\colrule
			Beam energy [TeV] & 0.98 & 6.5 & 7.0 & 7.0 & 50.0\\
			Number of bunches & 36 & 2220 & 2808 & 2748 & ? \\
			Number of particles per bunch & $2.90\times 10^{11}$ & $1.15\times 10^{11}$ & $1.15\times 10^{11}$ & $2.2\times 10^{11}$ & $1.0\times 10^{11}$\\
			Stored beam energy [MJ] & 1.6 & 265.9 & 362.2 & 678.0 & 8400 \\
		\end{tabular}
	\end{ruledtabular}
\end{table*}
Recent measurements at the LHC furthermore show that the tails of the transverse beam distribution are overpopulated compared to a Gaussian distribution resulting in a considerable amount of energy being stored in the beam tails alone. In case of the LHC explicitly around 5\% of the beam population is stored in the tails above 3.5~$\sigma$ compared to 0.22\% in case of a Gaussian distribution leading to 19~MJ of stored energy in the tails in case of nominal LHC parameters and 34~MJ in case of HL-LHC~\cite{helreview_valentino}. This leads to the conclusion that a mechanism is needed to deplete the beam tails in a controlled manner \cite{helreview}.

The most obvious idea is to decrease the collimator gaps or scrape the tails with a collimator type device. This approach is however not feasible as it would generate unacceptably large loss spikes due to the high intensity halo population in the LHC. Most promising are methods, which increase the diffusion speed in the region of the halo particles resulting in a smooth and continuous removal of the high amplitude tails while leaving the core of the beam unperturbed. The diffusing halo particles are then intercepted by the collimation system and removed. This concept is also referred to as active halo control compared to a passive system, which only intercepts the halo particles without actively controlling the diffusion speed (see Fig.~\ref{fig:active_halo_control} for illustration).
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.7\linewidth]{passive_halo_control.png}
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.7\linewidth]{active_halo_control.png}
	\end{minipage}	
	\caption{\label{fig:active_halo_control} Sketch of passive halo control as with the collimation system (top) and active halo control using in addition for example a hollow electron lens (HEL) to control the diffusion speed in the region tail region without affected the beam core (bottom).}
\end{figure}

In a recent review, the need for such an active halo control system for HL-LHC has been assessed with the conclusion that it would considerably increase the margin and reduce the risk for machine protection compared to the current LHC collimation system being a passive halo control system ~\cite{helreview}. In view of the need of active halo control for HL-LHC and for future high power accelerators like HE-LHC and FCC-hh~\cite{helhcparam2011,fcc_coll_ipac2017}, different active halo control methods have been studied in recent years~\cite{helreview_bruce}, of which the hollow electron lens (HEL) is considered the superior device for the HL-LHC~\cite{helreview}. The beneficial effect of a future HEL for HL-LHC in terms of machine protection and beam collimation can not be, however, deployed at the expenses of the beam core as latter would ultimately lead to a degradation of the performance due to increased particle losses and emittance growth. In this paper, we will concentrate on the assessment in simulations and measurements of possible detrimental effets on the beam core. We will summarize possible sources of perturbations of the proton beam core focusing in particular on the case of pulsed operation and the beam experiments at the LHC conducted in this context.

@@

\section{Hollow electron lens for HL-LHC\label{sec:hel}}
\subsection{General Overview\label{sec:hel:intro}}
Electron lenses produce DC or pulsed low energy electron beams, where the electron beam is generated in an electron gun, then guided and confined along a straight section by strong solenoids and finally dumped onto a collector. As an example the conceptual design of the HL-LHC HEL is shown in Fig.~\ref{fig:hel_layout}.
\begin{figure}[h]
	\includegraphics[width=1.0\linewidth]{hel_layout_epbeam}% Here is how to import EPS art
	\caption{\label{fig:hel_layout} Layout of HL-LHC HEL. Courtesy of CERN EN-MME group.}
\end{figure}

@@

			% 3-8 sigmap with 3.5 um -> 3.55-9.47 for 2.5 um emittance
			Desired range of scraping positions & 3.5-9.5 &$\sigma_p$\\

@@

\subsection{Operation modes and effects on the beam core\label{sec:hel:core}}
For the HEL two modes of operation are currently under consideration, the DC mode as standard operation mode and the pulsed mode. The main benefit of pulsed HEL operation is the increase in halo removal rates which might become essential as in ideal operation conditions featuring low non-linearities, in particular low chromaticity and octupole current, the DC mode is not effective~\cite{hel_halo_hllhc_fitterer,hl_halo_ipac2017}. 
For this purpose, two different pulsing patterns are thus considered for the HL-LHC:
\begin{itemize}
	\item \textbf{random excitation:} The electron gun voltage is modulated between:
	\begin{eqnarray}
	U_{\mathrm{e-gun}}&=&a\cdot U_{\mathrm{max}}\\
	& &+(1-a)\cdot \mathrm{ran}(0,1)\cdot U_{\mathrm{max}}
	\end{eqnarray}
	with $U_{\mathrm{max}}$ the equivalent voltage in DC operation, $a$ the modulation strength with $a\in[0,1]$, and $\mathrm{ran}(0,1)$ a uniformly distributed random number between~[0,1].
	\item \textbf{resonant excitation:} The HEL is switched on only every $k^{\mathrm{th}}$ turn. The excitation can then be represented by:
	\begin{eqnarray}\label{intro:eqn:1}
	f(t)&=&\sum_{p=-\infty}^{+\infty}\delta(t-n\cdot(kT)),
	\end{eqnarray}
	where $p$ is the turn number and $T$ the revolution time, and its Fourier series by:
	\begin{eqnarray}\label{intro:eqn:2}
	f(t)=\Sh_{kT}(t)
	&=&\frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt} \\
	& & \quad \text{with} \ f_n=\frac{n}{k}f_{\rm rev},
	\end{eqnarray}
	where $\Sh_{kT}$ is the Dirac comb. As can be seen from the Fourier series, $k^{\mathrm{th}}$~turn pulsing in general drives $k^{\mathrm{th}}$ order resonances \cite{md_sim_hel_res_ex_fitterer}. Historically, the $k^{\mathrm{th}}$~turn pulsing was used in regular operation in the Tevatron for abort cleaning \cite{hel_tevatron_abortgap_zhang}.
\end{itemize}

As depicted in Fig.~\ref{fig:hel_field} and in evidence from Eqs.~\ref{eq:field_1}--\ref{eq:field_2}, the field at the beam core vanishes in case of an ideal HEL. Effects on the beam core thus only arise from imperfections, where possible sources are the bends of the HEL (see Section~\ref{core:sec:1}), as here the electron beam crosses directly the proton beam, and distortions in the electron beam profile (see Section~\ref{core:sec:2}). Both sources result in non-linear kicks~\cite{hel_bends_stancari,hel_model_polynomial_morozov}. In DC operation, these non-linear kicks stay in the shadow of the non-linearities otherwise present in the machine if the electron beam is kept at a decent quality. Tolerances on imperfections are therefore not particularly stringent and of no concern. The picture changes significantly in case of pulsed operation, in which case the electron gun voltage is modulated using a white or colored noise spectrum. If the electromagnetic field of the HEL does not vanish at the proton beam core, noise is transferred not only to the halo particles, as intended, but also to the beam core. Tolerances on the residual field in this case rapidly become much more stringent than in the case of DC operation. Studies of the effect of the HEL on the beam core therefore focus on this mode of operation which will also be the main subject of this paper.

%In comparison, the random excitation mode is the most effective and operationally easiest pulsing mode, but also the most stringent in terms of tolerances on the residual field at the proton beam center as it induces white noise exciting all beam frequencies. As the resonant excitation acts on specific resonances, the pulsing pattern has to be chosen according to the machine and beam parameters. On the one hand, this makes the operation more difficult as one has to find the most efficient $k^{\mathrm{th}}$~turn pulsing for each operation mode, on the other hand, it also relaxes the tolerances on the residual field at the beam core as, depending on the machine and beam configuration, it might be possible to find a resonant excitation pattern, which mainly affects the halo and not the beam core.

@@

\subsection{Uncompensated kicks from HEL bends}
\label{core:sec:1}
As reference for the estimate of the dipole component originating from the HEL bends, the symplectic map derived in \cite{hel_bends_stancari} is used. In this case the HEL bends are modeled as a cylindrical pipe with a static charge distribution of 1~A and 5~keV electrons deflected in the horizontal plane, where the parameters were her still based on the old gun design for the Tevatron electron lens \textcolor{red}{Giulio, is this correct?}. In case of an U-shaped HEL, the transverse dipole kicks at entrance and exit add up, while for a S-shape the kicks compensate each other. For this reason a S-shape has been chosen for the HL-LHC HEL (see Fig.~\ref{fig:hel_layout}). Assuming and S-shaped HEL, uncompensated kicks therefore only arise from imperfections, which can originate from profile imperfections and current fluctuations. As a first estimate, we assume in this paper 10\% fluctuations between the entrance and exit and, in addition, that the kicks from entrance and exit due to these imperfections add up. Using the electric field calculations in \cite{hel_bends_stancari}, the maximum values for the integrated electric field are then
\begin{equation}
\int_{z_1}^{z_2} E_{x,y} dz= 10 \ \mathrm{kV}.
\end{equation}
%given an \SI{1}{A}, \SI{5}{keV} electron beam.
%For \SI{7}{TeV} protons and neglecting magnetic effects, this yields a kick of:
%\begin{equation}
%\Delta x'=\Delta y'= \SI{1.4}{nrad},
%\end{equation}
%Scaling to the HEL design parameters of a \SI{5}{A} and \SI{10}{keV} electron beam (Table~\ref{tab:hel_param}), one obtains:
%\begin{equation}
%\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
%\end{equation}
Neglecting magnetic effects and scaling to HL-LHC and HEL design parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}) yields a kick of \cite{md_sim_hel_res_ex_fitterer}:
\begin{equation}
\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
\end{equation}
Assuming now 10\% fluctuation between entrance and exit kick, the expected kick from the HEL bends is
\begin{equation}\label{eqn:kick_bends}
\Delta x', \Delta y'\approx \q{0.5}{nrad}.
\end{equation}

@@

\subsection{Kicks in the central region (main solenoid)}
\label{core:sec:2}
For a perfectly uniform, annular and radially symmetric profile, the field in the region of the proton beam core vanishes as can be seen from Eq.~\ref{eq:field_2} and is illustrated in Fig.~\ref{fig:hel_field}. Fields at the location of the proton beam core thus only arise in case of profile imperfections of the electron beam, in particular if the radial symmetry is broken. Recently, new measurements of the newest hollow electron gun CHG1b could be performed at the Fermilab electron lens test stand \cite{hel_test_stand_fnal}. A typical measurement of the electron beam current density obtained during these measurements is shown in Fig.~\ref{core:fig:0}.
\begin{figure}[t]
		\centering
		\includegraphics[width=0.8\linewidth]{e-field-measured.png}
	\caption{Current density from measurements of the CHG1b hollow electron gun at the Fermilab electron lens test stand in 2017 \cite{hel_res_field_stancari_2017}. \textcolor{red}{Replace by 2D simulations - Giulio.}}
	\label{core:fig:0}
\end{figure}
As at the test stand the solenoidal fields required for the HL-LHC electron lens can not be reached, an estimate of the electron beam profile for HL-LHC can only be obtained through simulations. For this purpose, the measured distribution is compressed to the inner electron beam radius of $0.97$~mm. Afterwards an input distribution with $40\ 000$ particles generated. The potential and fields can then be calculated with the particle in cell simulation code WARP~\cite{warp} yielding the relative electric field strengths shown in Fig.~\ref{core:fig:1}. The electric field can in general be scaled with the beam current and energy, but can not easily be scaled with the inner and outer radius of the electron beam distribution. Therefore it is essential to track a distribution with the correct dimensions. Further details on the measurements and WARP simulations can be found in~\cite{hel_res_field_stancari_2017}.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Emap.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{CHG1b_170512_8p75A_2-4-2p7kG_500V_76mA_hires_Eslice.png}
	\end{minipage}
	\caption{Calculated relative electric field for the new CHG1b hollow electron gun in the x-y plane and as projection on the x- and y-axis. The field calculations are based on measurements at the Fermilab test stand combined with WARP simulations for the calculation of the electric potentials and fields.}
	\label{core:fig:1}
\end{figure}

@@

\subsection{Overview of LHC experiments\label{sec:exp_sum}}
The experiments at the LHC which are presented in this paper were aiming at deriving tolerances on the residual HEL field at the location of the beam core in case of pulsed operation. In total, two experiments were conducted, the first one in 2016 \cite{resexmd2016} and the second one in 2017 \cite{resexmd2017}. Beam and machine parameters for both experiments are summarized in Table~\ref{tab:md_param}. Preceding the beam experiments, simulations were performed to identify the most effective $k^{\mathrm{th}}$~turn pulsing pattern and yield estimates for losses and emittance growth. The two $k^{\mathrm{th}}$~turn pulsing patterns with the largest effect, one $k^{\mathrm{th}}$~turn pulsing pattern with no effect and the random excitation could then be experimentally tested in the LHC. In order to test the reproducibility of the results at least on the example of one pulsing pattern, the $7^{\mathrm{th}}$~turn pulsing was first tried in 2016 and then repeated in 2017.

During the experiments the losses were measured with the fast beam current transformers (FBCTs) and the emittance and transverse beam profiles with the beam synchrotron radiation telescope (BSRT), where all instruments are capable of delivering bunch-by-bunch data. The analysis of the BSRT profiles is quite involved and we will therefore focus in this paper only on the direct comparison of the profiles. A detailed description of the analysis with more details can be found in \cite{bsrtprofinj} and for the individual experiments in \cite{resexmd2016,resexmd2017}.

\subsection{Excitation with transverse damper and filling scheme\label{sec:adt}}
The primary function of the transverse feedback system in LHC, also referenced to as ADT, is to provide injection oscillation damping and actively damp the coupled bunch instabilities driven by the machine impedance. The main building blocks of the ADT system are stipline pickups at the position Q7 and Q9 at Point~4 of the LHC, connected to the beam position measurement modules at the surface, the digital signal processing modules (mDSPU) and a set of tetrode power amplifiers feeding electrostatic kickers in the RF zone of the LHC~\cite{adt_sum_2008,adt_sum_2011}. Though originally designed to damp the undesired transverse activities only, the damper is being routinely used for sophisticated beam excitations as well thanks to the state of the art digital signal processing hardware. These include among others abort and injection gap cleaning, single bunch excitation for tune and linear coupling measurements and special modes of operation for dedicated experiments at the LHC.

The transverse feedback is in general active during the whole LHC cycle, therefore it is of very high importance, that the noise introduced by the system does not contribute to any emittance growth. The typical machine cycle requires a short damping time of ~10-20 turns (high feedback gain) for injection oscillation damping, which is increased during the ramp and collisions to ~50-100 turns (lower feedback gain). It has been demonstrated that noise introduced by the transverse feedback system itself is sufficiently low to introduce any measurable emittance growth \cite{adt_noise_emit_2017}. All excitation signals referenced to in this paper, are digitally synthesized in the damper's digital signal processing units and therefore can be considered as ``noise-free''. This point is important as then the effect on the beam can be attributed entirely to the applied excitation and any effect due to unwanted residual noise can be neglected.
